The Unicode Cookbook \linebreak for Linguists

### 2018-06-05 book/chapters/orthography_profiles.tex
deals with paperhivers comments for chapter 7 Orthography profiles

The major challenge, then, is developing an infrastructure to identify the
elements that are individual graphemes in a source, specifically for the
enormous variety of sources using some kind of alphabetic writing system.
Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
variety of writing systems that range from their own idiosyncratic
transcriptions to already well-established practical or longstanding
orthographies. Although the IPA is one practical choice as a sound-based
normalization for writing systems (which can act as an interlingual pivot to
attain interoperability across writing systems), graphemes in each writing
system must also be identified and standardized if interoperability across
different sources is to be achieved. In most cases, this amounts to more than
simply mapping a grapheme to an IPA segment because graphemes must first be
identified in context (e.g.~is the sequence one sound or two sounds or both?)
and strings must be tokenized, which may include taking orthographic rules into
account (e.g.~a nasal sound may be transcribed as /n/ when it appears between 
two vowels, but when it appears between a vowel and a consonant it becomes 
a nasalized vowel /Ṽ/).

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right} respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then they must be specified in the metadata file as such, 
	   and all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is 
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen such that they do not conflate
       with any symbols used in the orthography themselves. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, such as Unicode code points, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial, therefore it must be 
	   specified as so in the metadata file.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       then it should be made explicit in the orthography profile because regular 
	   expressions are executed in order from top to bottom.
\end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
        a warning might be given if the chosen string includes characters from
        the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
        tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
        proceed \textsc{global} or \textsc{linear} (see B6 above). 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
        should use \textsc{NFC} or \textsc{NFD}. 
    \end{enumerate}
	\item \textsc{might offer user-options}:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{to assist in the ordering of the graphemes.} In our experience, it
        makes sense to apply larger graphemes before shorter graphemes, and to
        apply graphemes with context before graphemes without context. Further,
        frequently relevant rules might be applied after rarely relevant rules
        (though frequency is difficult to establish in practice, as it depends
        on the available data). Also, if this all fails to give any decisive
        ordering between rules, it seems useful to offer linguists the option to
        reverse the ordering from any manual specified ordering, because
        linguists tend to write the more general rule first, before turning to
        exceptions or special cases. 
		\item \textsc{to assist in dealing with upper and lower case characters.} It
        seems practical to offer some basic case matching, so characters like
        <a> and <A> are treated equally. This will be useful in many concrete
        cases, although the user should be warned that case matching does not
        function universally in the same way across orthographies. Ideally,
        users should prepare orthography profiles with all lowercase and
        uppercase variants explicitly mentioned, so by default no case matching
        should be performed. 
		\item \textsc{to treat the profile literally}, i.e.~to not interpret regular
        expression metacharacters. Matching graphemes literally often leads to
        significant speed increase, and would allow users to not have to worry
        about escaping metacharacters. However, in our experience all actually
        interesting use-cases of orthography profiles include some contexts,
        which automatically prevents any literal interpretation.
    \end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings} to be processed in the specified
        Unicode normalization, i.e.~in either NFC or NFD as specified by the
        user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both in which
        text-strings any errors occurred and which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered. 
	\end{enumerate}
\end{enumerate}

>>>>

The major challenge, then, is developing an infrastructure to identify the
elements that are individual graphemes in a source, specifically for the
enormous variety of sources using some kind of alphabetic writing system.
Authors of source documents (e.g.~dictionaries, wordlists, corpora) use a
variety of writing systems that range from their own idiosyncratic
transcriptions to already well-established practical or longstanding
orthographies. Although the IPA is one practical choice as a sound-based
normalization for writing systems (which can act as an interlingual pivot to
attain interoperability across writing systems), graphemes in each writing
system must also be identified and standardized if interoperability across
different sources is to be achieved. In most cases, this amounts to more than
simply mapping a grapheme to an IPA segment because graphemes must first be
identified in context (e.g.~is the sequence one sound or two sounds or both?)
and strings must be tokenized, which may include taking orthographic rules into
account (e.g.~a nasal sound may be transcribed as <n> when it appears between 
two vowels, but when it appears between a vowel and a consonant it becomes 
a nasalized vowel <Ṽ>).

@@

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right} respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then they must be specified in the metadata file as such, 
	   and all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is 
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen so that they do not conflate
       with any symbols used in the orthography. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, such as Unicode code points, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression.} Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expression metacharacters in the profile.
       However, this is only possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial, therefore it must be 
	   specified as such in the metadata file.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       then it should be made explicit in the orthography profile because regular 
	   expressions are executed in order from top to bottom.
\end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{\arabic{enumi}.} 
	\item \textsc{should offer user-options} to specify:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} 
		\item \textsc{the name of the column to be used for transliteration} (if any). 
		\item \textsc{the symbol-string to be inserted between graphemes.} Optionally,
        a warning might be given if the chosen string includes characters from
        the orthography itself. 
		\item \textsc{the symbol-string to be inserted for unmatched strings} in the
        tokenized and transliterated output. 
		\item \textsc{the tokenization method}, i.e.~whether the tokenization should
        proceed as \textsc{global} or \textsc{linear} (see B6 above). 
		\item \textsc{unicode normalization}, i.e.~whether the text-string and profile
        should use \textsc{NFC} or \textsc{NFD}. 
    \end{enumerate}
	\item \textsc{might offer user-options}:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{5} 
		\item \textsc{to assist in the ordering of the graphemes.} In our experience 
		working with idiosyncratic transcriptions and orthographies from low-resource languages, it
        is helpful to identify multi-sequence graphemes before single graphemes, and to
        identify graphemes with context before graphemes without context. Further,
        frequently relevant rules might be applied after rarely relevant rules
        (though frequency is difficult to establish in practice, as it depends
        on the available data). Also, if this all fails to give any decisive
        ordering between rules, it seems useful to offer linguists the option to
        reverse the ordering from any manual specified ordering, because
        linguists tend to write the more general rule first, before turning to
        exceptions or special cases. 
		\item \textsc{to assist in dealing with upper and lower case characters.} It
        seems practical to offer some basic case matching, so characters like
        <a> and <A> are treated equally. This will be useful in many concrete
        cases (such as search or collation), although the user should be warned that case matching does not
        function universally in the same way across orthographies.\footnote{For example 
		compare the different first-letter capitalization practices of the digraphs 
		<ǋ> and <Ĳ> (single-character ligatures in the Unicode Standard) 
		in the Latin-based scripts of Southern-Slavic languages and 
		Dutch, respectively.} Ideally,
        users should prepare orthography profiles with all lowercase and
        uppercase variants explicitly mentioned, so by default no case matching
        should be performed. 
		\item \textsc{to treat the profile literally}, i.e.~to not interpret regular
        expression metacharacters. Matching graphemes literally often leads to
        significant speed increase, and ensures that users do not have to worry
        about escaping metacharacters. However, in our experience all actually
        interesting use-cases of orthography profiles include some contexts,
        which automatically prevents any literal interpretation.
    \end{enumerate}
	\item \textsc{should return the following information} to the user:
	\begin{enumerate}
		\def\labelenumii{C\arabic{enumii}.} \setcounter{enumii}{8} 
		\item \textsc{the original text-strings} to be processed in the specified
        Unicode normalization, i.e.~in either NFC or NFD as specified by the
        user. 
		\item \textsc{the tokenized strings}, with additionally any transliterated
        strings, if transliteration is requested. 
		\item \textsc{a survey of all errors encountered}, ideally both (i) in which
        text-strings any errors occurred and (ii) which characters in the
        text-strings lead to errors. 
		\item \textsc{a reordered profile}, when any automatic reordering is offered. 
	\end{enumerate}
\end{enumerate}
